A Comparison of Parallel Graph Processing Implementations

### 2017-05-10 papers/publication/paper-pollard.tex
some minor fixes to get paper to build for me, also started adding notes

@@

\begin{document}
\title{A Comparison of Parallel Graph Processing Implementations}
\author{
	\IEEEauthorblockN{Samuel D. Pollard}
	\IEEEauthorblockA{Computer and Information Science\\
		University of Oregon \\
		Eugene, OR 97403\\
		Email: spollard@cs.uoregon.edu}
	\and
	\IEEEauthorblockN{Boyana Norris}
	\IEEEauthorblockA{Computer and Information Science\\
		University of Oregon \\
		Eugene, OR 97403\\
		Email: norris@cs.uoregon.edu}
}
\maketitle
\begin{abstract}
The rapidly growing number of large network analysis problems has led to the emergence of many parallel and distributed graph processing systems---one survey in 2014 identified over 80. Since then, the landscape has evolved; some packages have become inactive while more are being developed. Determining the best approach for a given problem is infeasible for most developers. To enable easy, rigorous, and repeatable comparison of the capabilities of such systems, we present an approach and associated software for analyzing the performance and scalability of parallel, open-source graph libraries. We demonstrate our approach on five graph processing packages: GraphMat, the Graph500, the Graph Algorithm Platform Benchmark Suite, GraphBIG, and PowerGraph using synthetic and real-world\todo{actual real-world. (i) Michael LeBeane et al, Supercomputing (SC 15), Data Partitioning Strategies for Graph Workloads on Heterogeneous Clusters (ii) Shuang Song et al,  ICPP 2016, Proxy-Guided Load Balancing of Graph Processing Workloads on Heterogeneous Clusters.} datasets. We examine previously overlooked aspects of parallel graph processing performance such as phases of execution and energy usage for three algorithms\todo{more algorithms, justify why these algorithms}: breadth first search, single source shortest paths, and PageRank and compare our results to Graphalytics.
\end{abstract}

\section{Introduction}
%% What's the problem (motivation)
Our research is motivated by the current state of parallel graph processing. The most comprehensive survey, released in 2014, identified and categorized over 80 different parallel graph processing systems, not even considering domain specific languages \cite{Doekemeijer:2015:GPFSurvey}.

An overarching issue among these systems is the lack of comprehensive comparisons. One possible reason is the considerable effort involved in getting each system to run: satisfying dependencies and ensuring data are correctly formatted can be time consuming tasks. Additionally, some systems are developed with large, multi-node clusters in mind while others only work with shared memory, single node computers. An example of the former is GraphX \cite{Xin:2013:GraphX}, which incurs some overhead while gaining fault tolerance and an example of the latter is Ligra \cite{Shun:2013:Ligra}, a framework requiring a shared-memory architecture.
% ``The project was motivated by the fact that the largest publicly available real-world graphs all fit in shared memory''

From a graph algorithm user's point of view, optimizations for each system may not be apparent. For example, the Parallel Boost Graph Library (PBGL)~\cite{Gregor:2005:PBGL} provides generic implementations of their algorithms and the programmer must provide the template specializations. Optimal data structures may differ across graphs and must be determined by the programmer. To complicate matters further, the $\Delta$-stepping implementation of the parallel Dijkstra's Algorithm for computing single source shortest paths is parametrized by some $\Delta > 0$ for which requires tuning for a given architecture and graph structure to achieve optimal performance \cite{Panitanarak:2014:SSSPPerf}. \todo{Will this be mentioned again?}

@@

While leaderboards can indicate computational milestones, they are not particularly useful for the average user. Reference implementations, however, are critical for advancement; the complexity of today's hardware and operating systems require empirical evidence to assess an algorithm's performance. Additionally, there is motivation to define basic building blocks for graph computations \cite{GABB16, Buluc:CombBLAS:2011}. However, new algorithm designers and users alike have no easy way of determining what constitutes a high-performance implementation.

@@

\section{Related Work}
\todo{There's a large number of qualitative. Most third party comparisons are qualitative and just describe the overview instead of actual runtimes. This isn't a level playing field.}
Most performance analysis of graph processing systems come from the empirical results of each new library designer's publications. For example, GraphMat \cite{Sundaram:2015:GraphMat}, PowerGraph \cite{Gonzalez:2012:Powergraph}, and GraphBIG \cite{Nai:2015:Graphbig} present a comparison of performance between their system and a selection of other software packages.

@@

Beyond the aforementioned performance results there are a number of existing reports on performance analysis not tied to a particular software package we presently describe.

@@

The most prominent example of a graph processing performance analysis tool not tied to a particular iplementation is Graphalytics \cite{Capota:2015:Graphalytics}. Like our work, Graphalytics also automates the setup and execution of graph packages for performance analysis. Graphalyics relies on Apache Maven and Java to wrap each graph processing software package---Graphalytics calls these platforms---and requires configuration files for each dataset.\todo{What am I saying with this?}

@@

\section{Methodology}
\todo{Start with dataset subsection, write new intro to methodology, move this into inputs. We're comparing algorithms not applications. Understand those since they're the building blocks. Not comparing architectural details (cite that Nai paper). Important paragraph.}
Our utility is broken down into five principal phases indicated in \figurename~\ref{fig:epg-overview}, each of which requires no more than a single shell command. The five phases are as follows:
\begin{enumerate}
	\item Installing modified, stable forks of each software package to ensure homogeneity.
	\item Given a synthetic problem size or a real-world graph file, generate the files necessary to run each software package.
	\item Given a graph and the number of threads, run each algorithm using each software package multiple times.
	\item Parse through the log files to compress the output into a CSV.
	\item Analyze the data using the provided R scripts to generate plots.
\end{enumerate}

@@

\todo{More real world datasets}
The Graphalytics results in Table~\ref{tab:graphalytics} were performed on the Dota-League datasets. This dataset contains 61,670 vertices and 50,870,313 edges. This dataset comes from the Game Trace Archive\cite{Guo:2012:GTA} and is modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}. This dataset is useful because it is both weighted and more dense than the usual real-world dataset with an average out-degree of 824.

@@

This study explores four shared memory parallel graph processing platforms and one distributed memory framework. The first three are so-called ``reference implementations'' while the remaining two are included because of their performance and popularity. Other popular libraries such as the Parallel Boost Graph Library \cite{Gregor:2005:PBGL} are not considered here because the authors do not provide reference implementations. We consider the following systems.
\begin{itemize}
	\item The Graph500\footnote{We used the most recent version from \url{https://github.com/graph500/graph500}, most similar to release 2.1.4.} \cite{Murphy:2010:Graph500}, a set of large-scale graph benchmarks.
	\item The Graph Algorithm Platform (GAP) Benchmark Suite \cite{Beamer:2015:GAPBench}.
	\item GraphBIG \cite{Nai:2015:Graphbig} benchmark suite.
	\item GraphMat \cite{Sundaram:2015:GraphMat}, a high-performance graph library.
	\item PowerGraph \cite{Gonzalez:2012:Powergraph}, a framework for distributed graph-parallel computation on natural graphs.
	\todo{Explain each more.}
\end{itemize}

Our approach is not specific or limited to these graph packages and can be extended to others with relatively modest effort (building each package and if necessary, instrumenting the different phases of the computation).
\todo{Say what we automate and what we don't automate.}
\todo{Weighted versus unweighted}

@@

PageRank is selected because of its popularity; most libraries provide reference implementations. One challenge with using PageRank is the stopping criterion; all implementations have been modified to use $|| p_t - p_{t-1}||_1$ (the absolute sum of differences) where $p_t$ is the pagerank at step $t$. Verification of the PageRank results is beyond the scope of this paper, although this may explain some of the large performance discrepancies.\todo{Better explain why these were selected. They're the most popular of the algorithms for the papers we surveyed.}

@@

We analyze the performance of the algorithms described in Sec.~\ref{sec:algs} in terms of execution time, scalability over multiple threads, and power and energy consumption\todo{ A precise list and exposition of the requirements and changes	done to the application would help in understanding the methodology and its implications}. All figures except those measuring scalability (Fig.~\ref{fig:bfs-scaling}) use 32 threads.

@@

In Table~\ref{tab:graphalytics} we show the results from running Graphalytics on a Kronecker graph of scale $22$ (4,194,304 vertices and approximately $16 \times 2^22 = 33,500,000$ edges)\todo{work?}. An explanation of each algorithm is given in \cite{Iosup:2016:Graphalyticstech}. Graphalytics by default does not perform SSSP on unweighted, undirected graphs. The discrepancy between PageRank values in Table~\ref{tab:graphalytics} and Fig.~\ref{fig:pr} is a result of the differing stopping criterion.\todo{What are the likely culprits for performance differences---measure cache miss rates. Try to find something which correlates with performance. Resource stalls will work easily (not as interesting) Cycles per instruction? Automate with perfexplorer or autoperf}

@@

\todo{The y-axis is mislabeled. Should be "iterations" for pagerank.}
\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0pt 18pt 18pt 0pt, clip]{graphics/pr_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0pt, clip]{graphics/pr_iters.pdf}
	\end{minipage}
	\caption{The $y$-axis is logarithmic only for the left figure. GraphMat continues to run until none of the vertices' ranks change. For the others, we use the stopping condition that the sum of the changes in the weights is no more than $6 \times 10^{-8}$, or approximately machine epsilon for single precision floating point numbers.}
	\label{fig:pr}
\end{figure}

@@

\section{Future Work and Conclusion}
\todo{Maybe; it's still difficult to add more graphs to the dataset sinc you need to know all the metadata.}
We have presented a systematic approach to analyzing parallel graph processing performance, both measuring power consumption and the two fundamental phases of execution. While the comparison presented here can help one choose among alternatives for the selected packages and algorithms, the problem of selecting a graph processing framework for a given large-scale problem remains far from simple. Hence, we make our code available at \url{https://github.com/HPCL/easy-parallel-graph} and encourage further experimentation. Our results required minor changes to the original projects to add calls to PAPI sampling and those changes are also available on forked versions of the repositories.

@@

\bibliographystyle{IEEEtran}
\bibliography{../drp}
\end{document}

>>>>

\newcommand{\todob}[1]{\todo[inline,color=green!40]{#1}}

@@

\begin{document}
\title{A Comparison of Parallel Graph Processing Implementations}
\author{
	\IEEEauthorblockN{Samuel D. Pollard}
	\IEEEauthorblockA{Computer and Information Science\\
		University of Oregon \\
		Eugene, OR 97403\\
		Email: spollard@cs.uoregon.edu}
	\and
	\IEEEauthorblockN{Boyana Norris}
	\IEEEauthorblockA{Computer and Information Science\\
		University of Oregon \\
		Eugene, OR 97403\\
		Email: norris@cs.uoregon.edu}
}
\maketitle
\begin{abstract}
The rapidly growing number of large network analysis problems has led to the emergence of many parallel and distributed graph processing systems---one survey in 2014 identified over 80. Since then, the landscape has evolved; some packages have become inactive while more are being developed. Determining the best approach for a given problem is infeasible for most developers. To enable easy, rigorous, and repeatable comparison of the capabilities of such systems, we present an approach and associated software for analyzing the performance and scalability of parallel, open-source graph libraries. We demonstrate our approach on five graph processing packages: GraphMat, the Graph500, the Graph Algorithm Platform Benchmark Suite, GraphBIG, and PowerGraph using synthetic and real-world\todo[inline]{actual real-world. (i) Michael LeBeane et al, Supercomputing (SC 15), Data Partitioning Strategies for Graph Workloads on Heterogeneous Clusters (ii) Shuang Song et al,  ICPP 2016, Proxy-Guided Load Balancing of Graph Processing Workloads on Heterogeneous Clusters.} datasets. We examine previously overlooked aspects of parallel graph processing performance such as phases of execution and energy usage for three algorithms\todo[inline]{more algorithms, justify why these algorithms}: breadth first search, single source shortest paths, and PageRank and compare our results to Graphalytics.
\end{abstract}

\section{Introduction}
%% What's the problem (motivation)
\todob{need an introductory sentence about large graph applications and resulting proliferation of parallel frameworks}
Our research is motivated by the current state of parallel graph processing. The most comprehensive survey, released in 2014, identified and categorized over 80 different parallel graph processing systems, not even considering domain specific languages \cite{Doekemeijer:2015:GPFSurvey}.

An overarching issue among these systems is the lack of comprehensive comparisons. One possible reason is the considerable effort involved in getting each system to run: satisfying dependencies and ensuring data are correctly formatted can be time consuming tasks. Moreover, some systems are developed with large, multi-node clusters in mind while others only work with shared memory, single-node computers. An example of the former is GraphX \cite{Xin:2013:GraphX}, which incurs some overhead while gaining fault tolerance and an example of the latter is Ligra \cite{Shun:2013:Ligra}, a framework requiring a shared-memory architecture.
% ``The project was motivated by the fact that the largest publicly available real-world graphs all fit in shared memory''

From a graph algorithm user's point of view, optimizations for each system may not be apparent. For example, the Parallel Boost Graph Library (PBGL)~\cite{Gregor:2005:PBGL} provides generic implementations of their algorithms and the programmer must provide the template specializations. Optimal data structures may differ across graphs and must be determined by the programmer. To complicate matters further, the $\Delta$-stepping implementation of the parallel Dijkstra's Algorithm for computing single source shortest paths is parametrized by some $\Delta > 0$ for which requires tuning for a given architecture and graph structure to achieve optimal performance \cite{Panitanarak:2014:SSSPPerf}. \todo[inline]{Will this be mentioned again?}

@@

While leaderboards can indicate computational milestones, they are not particularly useful for the average user. Reference implementations, however, are critical for advancement; the complexity of today's hardware and operating systems requires empirical evidence to assess an algorithm's performance. Additionally, there is motivation to define basic building blocks for graph computations \cite{GABB16, Buluc:CombBLAS:2011}. However, new algorithm designers and users alike have no easy way of determining what constitutes a high-performance implementation.

@@

\section{Related Work}
\todo[inline]{There's a large number of qualitative. Most third party comparisons are qualitative and just describe the overview instead of actual runtimes. This isn't a level playing field.}
Most performance analysis of graph processing systems come from the empirical results of each new library designer's publications. For example, GraphMat \cite{Sundaram:2015:GraphMat}, PowerGraph \cite{Gonzalez:2012:Powergraph}, and GraphBIG \cite{Nai:2015:Graphbig} present a comparison of performance between their system and a selection of other software packages.

@@

Beyond the aforementioned performance results, a number of reports focus on performance analysis not tied to a particular software package we presently describe.

@@

The most prominent example of a graph processing performance analysis tool not tied to a particular iplementation is Graphalytics \cite{Capota:2015:Graphalytics}. Like our work, Graphalytics also automates the setup and execution of graph packages for performance analysis. Graphalyics relies on Apache Maven and Java to wrap each graph processing software package---Graphalytics calls these platforms---and requires configuration files for each dataset.\todo[inline]{What am I saying with this?}

@@

\section{Methodology}
\todo[inline]{Start with dataset subsection, write new intro to methodology, move this into inputs. We're comparing algorithms not applications. Understand those since they're the building blocks. Not comparing architectural details (cite that Nai paper). Important paragraph.}
Our utility is broken down into five principal phases indicated in \figurename~\ref{fig:epg-overview}, each of which requires no more than a single shell command. The five phases are as follows:
\begin{enumerate}
	\item Installing modified, stable forks of each software package to ensure homogeneity.
	\item Given a synthetic problem size or a real-world graph file, generate the files necessary to run each software package.
	\item Given a graph and the number of threads, run each algorithm using each software package multiple times.
	\item Parse through the log files to compress the output into a CSV.
	\item Analyze the data using the provided R scripts to generate plots.
\end{enumerate}

@@

\todo[inline]{More real world datasets}
The Graphalytics results in Table~\ref{tab:graphalytics} were performed on the Dota-League datasets. This dataset contains 61,670 vertices and 50,870,313 edges. This dataset comes from the Game Trace Archive\cite{Guo:2012:GTA} and is modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}. This dataset is useful because it is both weighted and more dense than the usual real-world dataset with an average out-degree of 824.

@@

This study explores four shared memory parallel graph processing platforms and one distributed memory framework. The first three are so-called ``reference implementations'' while the remaining two are included because of their performance and popularity. Other popular libraries such as the Parallel Boost Graph Library \cite{Gregor:2005:PBGL} are not considered here because the authors do not provide reference implementations. We consider the following systems.
\begin{itemize}
	\item The Graph500\footnote{We used the most recent version from \url{https://github.com/graph500/graph500}, most similar to release 2.1.4.} \cite{Murphy:2010:Graph500}, a set of large-scale graph benchmarks.
	\item The Graph Algorithm Platform (GAP) Benchmark Suite \cite{Beamer:2015:GAPBench}.
	\item GraphBIG \cite{Nai:2015:Graphbig} benchmark suite.
	\item GraphMat \cite{Sundaram:2015:GraphMat}, a high-performance graph library.
	\item PowerGraph \cite{Gonzalez:2012:Powergraph}, a framework for distributed graph-parallel computation on natural graphs.
	\todo[inline]{Explain each more.}
\end{itemize}

Our approach is not specific or limited to these graph packages and can be extended to others with relatively modest effort (building each package and if necessary, instrumenting the different phases of the computation).
\todo[inline]{Say what we automate and what we don't automate.}
\todo[inline]{Weighted versus unweighted}

@@

PageRank is selected because of its popularity; most libraries provide reference implementations. One challenge with using PageRank is the stopping criterion; all implementations have been modified to use $|| p_t - p_{t-1}||_1$ (the absolute sum of differences) where $p_t$ is the pagerank at step $t$. Verification of the PageRank results is beyond the scope of this paper, although this may explain some of the large performance discrepancies.\todo[inline]{Better explain why these were selected. They're the most popular of the algorithms for the papers we surveyed.}

@@

We analyze the performance of the algorithms described in Sec.~\ref{sec:algs} in terms of execution time, scalability over multiple threads, and power and energy consumption\todo[inline]{ A precise list and exposition of the requirements and changes	done to the application would help in understanding the methodology and its implications}. All figures except those measuring scalability (Fig.~\ref{fig:bfs-scaling}) use 32 threads.

@@

In Table~\ref{tab:graphalytics} we show the results from running Graphalytics on a Kronecker graph of scale $22$ (4,194,304 vertices and approximately $16 \times 2^22 = 33,500,000$ edges)\todo[inline]{work?}. An explanation of each algorithm is given in \cite{Iosup:2016:Graphalyticstech}. Graphalytics by default does not perform SSSP on unweighted, undirected graphs. The discrepancy between PageRank values in Table~\ref{tab:graphalytics} and Fig.~\ref{fig:pr} is a result of the differing stopping criterion.\todo[inline]{What are the likely culprits for performance differences---measure cache miss rates. Try to find something which correlates with performance. Resource stalls will work easily (not as interesting) Cycles per instruction? Automate with perfexplorer or autoperf}

@@

\todo[inline]{The y-axis is mislabeled. Should be "iterations" for pagerank.}
\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0pt 18pt 18pt 0pt, clip]{graphics/pr_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0pt, clip]{graphics/pr_iters.pdf}
	\end{minipage}
	\caption{The $y$-axis is logarithmic only for the left figure. GraphMat continues to run until none of the vertices' ranks change. For the others, we use the stopping condition that the sum of the changes in the weights is no more than $6 \times 10^{-8}$, or approximately machine epsilon for single precision floating point numbers.}
	\label{fig:pr}
\end{figure}

@@

\section{Future Work and Conclusion}
\todo[inline]{Maybe; it's still difficult to add more graphs to the dataset sinc you need to know all the metadata.}
We have presented a systematic approach to analyzing parallel graph processing performance, both measuring power consumption and the two fundamental phases of execution. While the comparison presented here can help one choose among alternatives for the selected packages and algorithms, the problem of selecting a graph processing framework for a given large-scale problem remains far from simple. Hence, we make our code available at \url{https://github.com/HPCL/easy-parallel-graph} and encourage further experimentation. Our results required minor changes to the original projects to add calls to PAPI sampling and those changes are also available on forked versions of the repositories.

@@

\bibliographystyle{IEEEtranS}
\bibliography{../drp}
\end{document}
